Assessment and optimization of respiratory syncytial virus prophylaxis regimens in Connecticut, \Sexpr{formatYear(startDate)}-\Sexpr{formatYear(endDate)}

### 2019-02-15 Paper/RSVOptimization.rnw
Single-column

% TODO: Plot regression line in onset/offset analysis
% TODO clarify all years analysis in figure captions
% TODO Put 1-2% punchline in the abstract, maybe
% TODO So it turns out that several regimens are actually the same, and that the best regional regimen @ 2w rounding is actually the same as the best statewide regimen @ 2w rounding. As a result, the conclusion could be simplified to statewide 2w rounding.
% TODO paragraphs

@@

<<load>>=
read_chunk("../Analysis/RSV.R")
read_chunk("../Thesis/Thesis.R")
@

@@

\author{
    \authorstyle{Ben Artin\textsuperscript{1}}
    \newline\newline
    \authorstyle{Daniel Weinberger, PhD\textsuperscript{2}, advisor}\\
    \authorstyle{Virginia Pitzer, ScD\textsuperscript{2}, advisor}
    \newline\newline
    \textsuperscript{1}\institution{Yale University, School of Public Health and School of Medicine}\\
    \textsuperscript{2}\institution{Yale University, School of Public Health}
}

@@

\begin{center}

\vspace{20pt}
\large Submitted to the School of Public Health and the School of Medicine, in partial fulfillment of the requirements for the degrees of Master of Public Health and Master of Medical Sciences

\vspace{40pt}

\includegraphics[height=160pt]{EPH}
\hspace{40pt}
\includegraphics[height=160pt]{Medicine}

\end{center}

@@

\cleardoublepage
\endgroup

@@

\abstractsection{Background}Respiratory syncytial virus (RSV) causes seasonal respiratory infection with potentially serious complications in children, and leads to hospitalization rates as high as 50\% in high-risk infants. Safe and effective immunoprophylaxis (palivizumab) is available, but costly. The American Academy of Pediatrics (AAP) recommends palivizumab only for infants at high risk of complications, and only during the RSV season. Although the current AAP guidelines acknowledge the existence of spatial and temporal variation in RSV incidence, they do not recommend spatial or temporal adjustments to immunoprophylaxis regimen outside of Florida. In this study, we investigate the value of using spatial and temporal variation in RSV incidence to adjust the RSV prophylaxis regimen in Connecticut.

@@

\abstractsection{Results}Using preventable fraction to assess different immunoprophylaxis regimens, we found that regimens adjusted for county-level variation in timing of RSV seasons are superior to the current AAP-recommended regimen. We also considered the effect of rounding the timing of the first dose of prophylaxis to pragmatic calendrical boundaries (weekly, biweekly, and monthly), and found that benefits of alternative regimens over the AAP-recommended regimen persisted with biweekly rounding, but not with monthly rounding. Our best-performing pragmatic alternative to the AAP guidelines was based on the regional RSV season midpoint with biweekly rounding. In comparison to the AAP recommendation, whose preventable fraction is \Sexpr{regionStrategyCoverage("all", "aap", 0)} (\Sexpr{regionStrategyCoverageCI("all", "aap", 0, level)}), that alternative yielded improvement to preventable fraction of \Sexpr{regionStrategyCoverage("all", "countyMiddle", 2)} (\Sexpr{regionStrategyCoverageCI("all", "countyMiddle", 2, level)}). We also found that alternative regimens adjusted for annual variation in RSV season are non-superior to spatially adjusted regimens.

@@

\cleardoublepage

\pagestyle{fancy}
\tableofcontents
\clearpage

@@

\begin{multicols}{2}

@@

For infants, risk factors of serious illness and hospitalization due to RSV infection include prematurity, chronic lung disease of prematurity, congenital heart disease, anatomic pulmonary abnormalities, neuromuscular disorders, trisomy 21, and immunocompromised status.\cite{AmericanAcademyofPediatricsCommitteeonInfectiousDiseases:2014bj} Hospitalization rate due to RSV among high-risk infants is anywhere from two to ten times higher than among infants with no comorbidities.\cite{Boyce:2000gz}

@@

Palivizumab has been found to be effective in reducing hospitalizations due to RSV; in double-blinded trials in high-risk infants, it reduced hospitalization rate by 45-55\% compared to placebo.\cite{Group:1998ih,Feltes:2003wm} Its has also been found to be safe and well-tolerated in clinical trials.\cite{Group:1998ih}

Owing to the high cost, the American Academy of Pediatrics (AAP) guidelines, revised in 2014, recommend palivizumab immunoprophylaxis only for high-risk infants, and only during RSV season.\cite{Anonymous:1998up,AmericanAcademyofPediatricsCommitteeonInfectiousDiseasesandCommitteeonFetusandNewborn:2003ug,AmericanAcademyofPediatricsCommitteeonInfectiousDiseases:2014bj}

@@

RSV seasonality has been established throughout the United States, based on National Respiratory and Enteric Virus Surveillance System data. In all states except for Florida, it shows a stable season onset between mid-November and early January and a stable season offset between mid-March and late April.\cite{CentersforDiseaseControlandPreventionCDC:2011wt,CentersforDiseaseControlandPreventionCDC:2013wv} RSV season in Florida begins approximately two months earlier than in other states. \cite{CentersforDiseaseControlandPreventionCDC:2013wv}

The AAP acknowledges the potential significance of spatial and temporal variation in RSV incidence, but --- outside of Florida --- does not recommend local variance from nation-wide prophylaxis guidelines.\cite{AmericanAcademyofPediatricsCommitteeonInfectiousDiseases:2014bj}

@@

Nationwide spatial variability of RSV season has been established by the Centers for Disease Control and Prevention (CDC).\cite{CentersforDiseaseControlandPreventionCDC:2011wt,CentersforDiseaseControlandPreventionCDC:2013wv}

In Connecticut, the Department of Public Health (CT DPH), via its Office of Health Care Access (OHCA), maintains a database of all hospitalizations in every non-federal short-stay acute-care general hospital in the state. Previous analysis of this database at the Yale University School of Public Health has shown regional variability in time of peak RSV incidence within Connecticut.\cite{Noveroske:y4fi3188} No prior research is available regarding year-to-year variation in timing of RSV season in Connecticut.

@@

\begin{itemize}
	\item To what extent does regional variability of RSV incidence in Connecticut impact the effectiveness of the AAP-recommended prophylaxis regimen?
	\item Can we use spatial analysis of RSV to propose alternative regimens that would increase the benefits of prophylaxis in comparison to the AAP-recommended regimen, without increasing the total cost of the prophylactics the regimen?
	\item Is there year-to-year variation in timing of RSV season in Connecticut?
	\item Can we take advantage of any year-to-year variation to improve upon AAP-recommended prophylaxis, without increasing treatment cost?
\end{itemize}

@@

\begin{itemize}
    \item the postal code of residence listed in patient's medical record was in Connecticut\footnote{According to the zipcode $\rightarrow$ state mapping in R \texttt{zipcode} package version 1.0}, and
    \item the hospital stay was associated with ICD-9-CM diagnosis code \texttt{079.6} (RSV), \texttt{466.11} (Acute bronchiolitis due to RSV), or \texttt{480.1} (Pneumonia due to RSV).
\end{itemize}

Using the OHCA weekly hospitalization statistics, we identified \Sexpr{nrow(dataset)} RSV-associated hospitalizations in Connecticut between \Sexpr{formatMonth(startDate)} and \Sexpr{formatMonth(endDate)} (inclusive).

@@

Given that RSV seasons peak early in each calendar year, we assigned to each surveillance year 12 months of data starting in July of previous year. OHCA provided us with hospitalization records aggregated into Monday-Sunday calendar weeks; we assigned the first full calendar week of July to be surveillance week 1 in each surveillance year.

Due to changes in diagnostic coding, we excluded data prior to July 1996; we similarly excluded data starting after June 2013, due to incompleteness.

@@

We were willing to assume that RSV incidence is the result of a Poisson process whose $\log$ can be modeled as a piecewise twice-differentiable function of time. To meet those criteria and assumptions, we chose a log-linked generalized additive model (GAM) with penalized cubic B-splines for our analysis.

@@

We chose to aggregate low-RSV counties into a single region because our analysis in those counties individually would have yielded results with unacceptably low statistical power.

@@

We began by assessing the current AAP RSV prophylaxis guidelines. These guidelines recommend RSV prophylaxis with 5 monthly doses of palivizumab, starting in November, for all high-risk infants in their first year of life; in the second year of life, only some subgroups are recommended to receive palivizumab.

For simplicity, we assumed that protection by palivizumab begins on November 15th and lasts 24 weeks.

@@

The choice of \Sexpr{seasonThreshold * 100}\% and \Sexpr{(1 - seasonThreshold) * 100}\% incidence thresholds was based on our intent to use estimates of RSV season onset and offset to put forth alternative prophylaxis regimens (as described below), while maintaining 5-month dosing schedule (and therefore not increasing medication cost). Incidence cutoffs other than \Sexpr{seasonThreshold * 100}\% and \Sexpr{(1 - seasonThreshold) * 100}\% were briefly investigated, but quickly ruled out because they produced prophylaxis regimens starting long prior to RSV season onset or ending long after RSV season offset, and therefore had very low preventable fractions.

@@

Our analysis thus far assumed that RSV seasons are similar year-to-year. To analyze the effects of year-to-year variation between RSV seasons on prophylaxis regimens, we chose to aggregate our data into (overlapping) periods of three consecutive years --- in order to maintain a higher level of statistical power, as well as to limit impact of short-term variation on our results.

This gave us \Sexpr{year(endDate) - year(startDate) -1} three-year periods, with the \Sexpr{formatYear(startDate + 2*365)} period covering surveillance years \Sexpr{formatYear(startDate)} through \Sexpr{formatYear(startDate + 2*365)}, and the \Sexpr{formatYear(endDate)} period covering \Sexpr{formatYear(endDate - 2*365)} through \Sexpr{formatYear(endDate)}.

We began our temporal analysis by estimating onset and offset for each three-year period, using the same penalized spline GAM as all-years analysis, producing --- as before --- statewide and regional estimates of \Sexpr{seasonThreshold * 100}\% onset and \Sexpr{(1 - seasonThreshold) * 100}\% offset.

@@

Meanwhile, in comparison with the statewide RSV season onset and offset, the low-RSV region has RSV season with later onset at \Sexpr{regionOnset("lowIncidence")} weeks (\Sexpr{regionOnsetCI("all", level)} weeks); season offset shows no change from the statewide median (Figure \ref{fig:seasonByRegion}).

@@

This is caused by temporal misalignment between the RSV season (delineated by median \Sexpr{seasonThreshold * 100}\%  onset and \Sexpr{(1 - seasonThreshold) * 100}\% offset in each region), and the timing of the AAP-recommended prophylaxis regimen (Figure \ref{fig:coverageByRegionAAPInsight}). Fairfield county, owing to its early season onset, sees disproportionately many cases before prophylaxis administration begins; on the other hand, the low-RSV counties, owing to their late season onset and relatively shorter RSV seasons, see almost all of their RSV cases within the prophylaxis window.

@@

\columnbreak

>>>>

@@

<<load>>=
read_chunk("../Analysis/RSV.R")
read_chunk("../Paper/RSVOptimization.R")
@

@@

\author{
    \authorstyle{Ben Artin, MEng, MPH, MMSc\textsuperscript{1}}
    \newline\newline
    \authorstyle{Daniel Weinberger, PhD\textsuperscript{1}}\\
    \authorstyle{Virginia Pitzer, ScD\textsuperscript{1}}
    \newline\newline
    \textsuperscript{1}\institution{Yale University, School of Public Health}
}

@@

@@

\clearpage
\endgroup

@@

\abstractsection{Background}Respiratory syncytial virus (RSV) causes seasonal respiratory infection with potentially serious complications in children, and leads to hospitalization rates of up to 50\% in high-risk infants. Safe and effective immunoprophylaxis (palivizumab) is available, but costly. The American Academy of Pediatrics (AAP) recommends palivizumab only for infants at high risk of complications, and only during the RSV season. Although the current AAP guidelines acknowledge the existence of spatial and temporal variation in RSV incidence, they do not recommend spatial or temporal adjustments to the immunoprophylaxis regimen outside of Florida. In this study, we investigate the value of using spatial and temporal variation in RSV incidence to adjust the RSV prophylaxis regimen in Connecticut.

@@

\abstractsection{Results}Using preventable fraction to assess different immunoprophylaxis regimens, we found that regimens adjusted for county-level variation in timing of RSV seasons are superior to the current AAP-recommended regimen. We also considered the effect of rounding the timing of the first dose of prophylaxis to pragmatic calendrical boundaries (weekly, biweekly, and monthly), and found that benefits of alternative regimens over the AAP-recommended regimen persisted with biweekly rounding, but not with monthly rounding. Our best-performing pragmatic alternative to the AAP guidelines was based on the regional RSV season midpoint with biweekly rounding. In comparison to the AAP recommendation, whose preventable fraction is \Sexpr{regionStrategyCoverage("all", "aap", 0)} (\Sexpr{regionStrategyCoverageCI("all", "aap", 0, level)}), that alternative yielded improvement to \Sexpr{regionStrategyCoverage("all", "countyMiddle", 2)} (\Sexpr{regionStrategyCoverageCI("all", "countyMiddle", 2, level)}). We also found that alternative regimens adjusted for annual variation in the RSV season are non-superior to spatially adjusted regimens.

@@

\clearpage

\pagestyle{fancy}

@@

%\begin{multicols}{1}

@@

For infants, risk factors of serious illness and hospitalization due to RSV infection include prematurity, chronic lung disease of prematurity, congenital heart disease, anatomic pulmonary abnormalities, neuromuscular disorders, trisomy 21, and immunocompromised status.\cite{AmericanAcademyofPediatricsCommitteeonInfectiousDiseases:2014bj} The hospitalization rate due to RSV among high-risk infants is anywhere from two to ten times higher than among infants with no comorbidities.\cite{Boyce:2000gz}

@@

Palivizumab has been found to be effective in reducing hospitalizations due to RSV; in double-blinded trials in high-risk infants, it reduced the hospitalization rate by 45-55\% compared to placebo.\cite{Group:1998ih,Feltes:2003wm} Its has also been found to be safe and well-tolerated in clinical trials.\cite{Group:1998ih}

Owing to the high cost, the American Academy of Pediatrics (AAP) guidelines, revised in 2014, recommend palivizumab immunoprophylaxis only for high-risk infants, and only during the RSV season.\cite{Anonymous:1998up,AmericanAcademyofPediatricsCommitteeonInfectiousDiseasesandCommitteeonFetusandNewborn:2003ug,AmericanAcademyofPediatricsCommitteeonInfectiousDiseases:2014bj}

@@

RSV seasonality has been established throughout the United States, based on National Respiratory and Enteric Virus Surveillance System data. In all states except for Florida, it shows an annual or biennial cycle with a stable seasonal onset between mid-November and early January and a stable seasonal offset between mid-March and late April.\cite{CentersforDiseaseControlandPreventionCDC:2011wt,CentersforDiseaseControlandPreventionCDC:2013wv} RSV season in Florida begins approximately two months earlier than in other states. \cite{CentersforDiseaseControlandPreventionCDC:2013wv}

The AAP acknowledges the potential significance of spatial and temporal variation in RSV incidence, but --- outside of Florida --- does not recommend local variance from nationwide prophylaxis guidelines.\cite{AmericanAcademyofPediatricsCommitteeonInfectiousDiseases:2014bj}

@@

TODO Nationwide spatial variability of RSV season has been established by the Centers for Disease Control and Prevention (CDC).\cite{CentersforDiseaseControlandPreventionCDC:2011wt,CentersforDiseaseControlandPreventionCDC:2013wv}

In Connecticut, the Department of Public Health (CT DPH), via its Office of Health Care Access (OHCA), maintains a database of all hospitalizations in every non-federal short-stay acute-care general hospital in the state. Previous analysis of this database has shown regional variability in time of peak RSV incidence within Connecticut.\cite{Noveroske:y4fi3188} No prior research is available regarding year-to-year variation in timing of RSV season in Connecticut.

@@

\begin{itemize}
	\item To what extent does regional variability of RSV incidence in Connecticut impact the effectiveness of the AAP-recommended prophylaxis regimen?
	\item Can we use spatial analysis of RSV to propose alternative regimens that would increase the benefits of prophylaxis in comparison to the AAP-recommended regimen, without increasing the total cost of the regimen?
	\item Is there year-to-year variation in timing of RSV season in Connecticut?
	\item Can we take advantage of any year-to-year variation to improve upon AAP-recommended prophylaxis, without increasing treatment cost?
\end{itemize}

@@

\begin{itemize}
    \item the postal code of residence listed the in patient's medical record was in Connecticut, and
    \item the hospital stay was associated with ICD-9-CM diagnosis code \texttt{079.6} (RSV), \texttt{466.11} (Acute bronchiolitis due to RSV), or \texttt{480.1} (Pneumonia due to RSV).
\end{itemize}

Using the OHCA weekly hospitalization statistics, we identified \Sexpr{prettyNum(nrow(dataset), big.mark=",")} RSV-associated hospitalizations in Connecticut between \Sexpr{formatMonth(startDate)} and \Sexpr{formatMonth(endDate)} (inclusive).

@@

Given that RSV seasons peak early in each calendar year, we designated the surveillance year as the 12-month period beginning  in July. OHCA provided us with hospitalization records aggregated into Monday-Sunday calendar weeks; we assigned the first full calendar week of July to be surveillance week 1 in each surveillance year.

Due to changes in diagnostic coding, we excluded data prior to July 1996; we similarly excluded data starting after June 2013 due to incompleteness.

@@

We assumed that RSV incidence is the result of a Poisson process whose $\log$ can be modeled as a piecewise twice-differentiable function of time. To meet those criteria and assumptions, we chose a log-linked generalized additive model (GAM) with penalized cubic B-splines for our analysis.

@@

We chose to aggregate low-RSV counties into a single region because our analysis in those counties individually would have yielded results with low statistical power.

@@

We began by assessing the current AAP RSV prophylaxis guidelines. These guidelines recommend RSV prophylaxis with 5 monthly doses of palivizumab, starting in November, for all high-risk infants in their first year of life; in the second year of life, palivizumab is only recommended for some subgroups. For simplicity, we assumed that protection by palivizumab begins on November 15th and lasts 24 weeks.

@@

The choice of \Sexpr{seasonThreshold * 100}\% and \Sexpr{(1 - seasonThreshold) * 100}\% incidence thresholds was based on our intent to use estimates of RSV season onset and offset to put forth alternative prophylaxis regimens (as described below), while maintaining the 5-month dosing schedule (and therefore not increasing medication cost). Incidence cutoffs other than \Sexpr{seasonThreshold * 100}\% and \Sexpr{(1 - seasonThreshold) * 100}\% were briefly investigated, but quickly ruled out because they produced prophylaxis regimens starting long prior to RSV season onset or ending long after RSV season offset, and therefore had very low preventable fractions.

@@

Our analysis thus far assumed that RSV seasons are similar year-to-year. To analyze the effects on prophylaxis regimens of year-to-year variation between RSV seasons, we chose to aggregate our data into (overlapping) periods of three consecutive years --- in order to maintain a higher level of statistical power, as well as to limit the impact of short-term variation on our results.

TODO This gave us \Sexpr{year(endDate) - year(startDate) -1} three-year periods, with the \Sexpr{formatYear(startDate + 2*365)} period covering surveillance years \Sexpr{formatYear(startDate)} through \Sexpr{formatYear(startDate + 2*365)}, and the \Sexpr{formatYear(endDate)} period covering \Sexpr{formatYear(endDate - 2*365)} through \Sexpr{formatYear(endDate)}.

We began our temporal analysis by estimating onset and offset for each three-year period, using the same penalized spline GAM as for the all-years analysis, producing --- as before --- statewide and regional estimates of \Sexpr{seasonThreshold * 100}\% onset and \Sexpr{(1 - seasonThreshold) * 100}\% offset.

@@

TODO Meanwhile, in comparison with the statewide RSV season onset and offset, the low-RSV region has RSV season with later onset at \Sexpr{regionOnset("lowIncidence")} weeks (\Sexpr{regionOnsetCI("all", level)} weeks); season offset shows no change from the statewide median (Figure \ref{fig:seasonByRegion}).

@@

This is caused by temporal misalignment between the RSV season (delineated by median \Sexpr{seasonThreshold * 100}\%  onset and \Sexpr{(1 - seasonThreshold) * 100}\% offset in each region) and the timing of the AAP-recommended prophylaxis regimen (Figure \ref{fig:coverageByRegionAAPInsight}). Fairfield county, owing to its early season onset, sees disproportionately more cases before prophylaxis administration begins; on the other hand, the low-RSV counties, owing to their late season onset and relatively shorter RSV seasons, see almost all of their RSV cases within the prophylaxis window.

@@

%\columnbreak
